---
title: "Conventions for handling negative response values: simulation results"
format:
  html:
    toc: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(ggplot2); library(dplyr); library(tidyr)

# Everything the report draws comes from the tables analysis/collate.R writes,
# so a refresh changes the numbers without touching prose. Each is optional:
# the 24 case units finish long before the 4,200 simulation units, so the report
# has to render on whichever half exists rather than failing on the other.
read_if <- function(...) {
  f <- here::here("results", ...)
  if (file.exists(f)) read.csv(f) else NULL
}
m          <- read_if("metrics.csv")
weights    <- read_if("weights.csv")
diagnostics <- read_if("diagnostics.csv")
cases      <- read_if("case_estimates.csv")
case_diag  <- read_if("case_diagnostics.csv")
have_sim   <- !is.null(m)
have_cases <- !is.null(cases)
if (!have_sim && !have_cases) stop("no collated results; run analysis/collate.R")

ARMS <- c("measured", "censored", "deleted", "floored", "beta", "gamma")
# Hue order validated for colour-vision separation as a categorical set: every
# adjacent pair clears the CVD and normal-vision floors on a light surface.
# Reordering the arms means revalidating, because the check is on adjacent pairs.
ARM_COL <- c(measured = "#1B6BC4", censored = "#0E7C61", deleted  = "#C25E12",
             floored  = "#B0217A", beta     = "#7C7A00", gamma    = "#6B4EBF")
EST <- c("ErC10", "ErC50", "NSEC")
if (have_sim) {
  m$arm <- factor(m$arm, levels = ARMS)
  m$estimate <- factor(m$estimate, levels = EST)
  prec <- m |>
    filter(cell %in% c("p1", "p2", "p3", "p4")) |>
    mutate(cv = factor(sprintf("%.1f", 100 * cv_control),
                       levels = c("9.6", "6.7", "2.8", "1.9")))
}
if (have_cases) {
  cases$arm <- factor(cases$arm, levels = ARMS)
  cases$estimate <- factor(cases$estimate, levels = EST)
}

# Read a figure out of the table rather than typing it into the prose, so a
# re-run cannot leave the text describing an earlier one. The previous version
# of this report had every number in "What the study establishes" typed by hand,
# and every one of them was superseded by the re-run.
val <- function(a, est, cl, what = "rel_bias") {
  if (!have_sim) return(NA_real_)
  m[[what]][m$arm == a & m$estimate == est & m$cell == cl]
}
# An em-dash rather than "NA" where the half this figure comes from has not been
# collated yet, so a report rendered mid-run reads as incomplete rather than
# broken.
f1 <- function(x) {
  if (!length(x) || any(is.na(x))) return("\u2014")
  sub("^-0\\.0$", "0.0", sprintf("%.1f", x))
}
f2 <- function(x) {
  if (!length(x) || any(is.na(x))) return("\u2014")
  sprintf("%.2f", x)
}
# The spread between the four Gaussian conventions in one cell, in percentage
# points: what a reader deciding whether the choice matters actually needs.
GAUSS <- c("measured", "censored", "deleted", "floored")
spread <- function(est, cl) {
  if (!have_sim) return(NA_real_)
  v <- vapply(GAUSS, val, numeric(1), est, cl)
  max(v) - min(v)
}

# One label per series at the right-hand edge instead of a key: six hues in a
# legend are harder to match to lines than a label sitting on the line it names.
edge_label <- function(d, yvar) {
  geom_text(data = subset(d, cv == "1.9"),
            aes(label = arm, y = .data[[yvar]]), hjust = -0.15, size = 2.6,
            show.legend = FALSE)
}
```

## The study

We simulated concentration-response data from a four-parameter curve
(`nec4param`) in which specific growth rate falls from a positive control value
to a negative lower asymptote, and we handed each simulated dataset to six
analyses that differ only in how they treat the negative values. Every analysis
uses `bayesnec` with package defaults, model-averaged over the declining
candidate set the family allows, so a difference between arms is the difference
a practitioner would obtain rather than a likelihood-only contrast.

```{r arms-table}
knitr::kable(
  data.frame(
    arm = ARMS,
    `what the analyst does` = c(
      "use the values as recorded",
      "declare negatives left-censored at zero",
      "delete the negative rows, keep the rest of their group",
      "replace negatives with zero",
      "floor, scale to (0, 1), fit a Beta",
      "floor, fit a Gamma"),
    family = c("gaussian", "gaussian", "gaussian", "gaussian", "Beta", "Gamma"),
    check.names = FALSE),
  caption = "**Table 1.** The six conventions compared. All six are reachable from `bnec()` defaults.")
```

Seven design cells were run, each 100 times. Four vary measurement precision
alone, holding the design and the true values fixed, so that the control
coefficient of variation falls from 9.6% to 1.9%. Two vary the depth of the
lower asymptote. One stops the concentration series at the point where the true
curve reaches zero growth, so that almost nothing is negative and the
conventions have nothing to act on.

The true values are known by construction, so each analysis is scored against
them rather than against each other. We report relative bias, the proportion of
95% credible intervals containing the truth, and the mean interval width.

Four real datasets are analysed by the same six conventions as well. They have
no known truth, so they are not scored; what they show is how far the choice
changes the number that would be reported on data nobody simulated.

Everything here comes from one pinned commit of `bayesnec`, and the results it
replaces came from another. The Provenance section says which, and what changed
between them.

## Estimating ErC50

The concentration halving the growth rate separates the conventions cleanly, and
it separates them by whether the measurement is kept.

```{r fig-erc50-bias, fig.height=4.2}
#| eval: !expr have_sim
d <- filter(prec, estimate == "ErC50")
ggplot(d, aes(cv, rel_bias, colour = arm, group = arm)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "rel_bias") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Relative bias in ErC50 (%)") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 1.** Relative bias in the estimated ErC50 against measurement
precision, 100 simulated datasets per point. Arms that keep the measurement
converge on the true value; arms that replace it level off away from zero.

Three arms converge as the experiment improves: `measured` from -0.8% to -0.2%,
`censored` from -1.3% to -0.2%, and `deleted` from -2.5% to 0.0%. Their error is
estimation error, and it behaves like estimation error.

Two do not. `floored` settles at about -2.3% and `beta` at about -6.0%, and
neither improves as the control coefficient of variation falls from 9.6% to
1.9%. A discrepancy that does not shrink as the noise shrinks is
misspecification, not imprecision, and no amount of experimental effort removes
it.

`gamma` is a special case that the bias alone misrepresents. Its ErC50 bias is
near zero in every cell, but its intervals are more than twice as wide as any
other arm's and its ErC10 is badly displaced, as the next section shows.

## Estimating ErC10

```{r fig-erc10-bias, fig.height=4.2}
#| eval: !expr have_sim
d <- filter(prec, estimate == "ErC10")
ggplot(d, aes(cv, rel_bias, colour = arm, group = arm)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "rel_bias") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Relative bias in ErC10 (%)") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 2.** Relative bias in the estimated ErC10. `gamma` is displaced by
about 20% at every level of precision.

The four Gaussian arms are within a few per cent of the truth once the
measurement is precise enough to resolve a 10% decline, and they converge.
`gamma` does not: it underestimates ErC10 by 19% to 24% in every cell of the
precision axis, and the error is flat, so a better experiment does not help.
`beta` drifts the other way, from -4.8% to +8.9%.

ErC10 is the harder estimate for every arm at realistic noise, because a 10%
decline from the control has to be resolved against a control coefficient of
variation of 9.6%. That is a property of the endpoint rather than of any
convention, and it is why the precision axis is necessary to interpret it.

## The NSEC

The no-significant-effect concentration is where the conventions diverge most,
and where they diverge in a direction that matters for protection.

```{r fig-nsec-bias, fig.height=4.2}
#| eval: !expr have_sim
d <- filter(prec, estimate == "NSEC")
ggplot(d, aes(cv, rel_bias, colour = arm, group = arm)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "rel_bias") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Relative bias in the NSEC (%)") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 3.** Relative bias in the estimated NSEC against the true no-effect
concentration. The three arms that keep the measurement converge toward it from
above; the three that replace it diverge downward as precision improves.

The NSEC's expected value exceeds the true `nec` at any finite level of noise,
because it is defined against a lower quantile of the control posterior: a
noisier control gives a lower reference and pushes the crossing to the right. Its
limiting value as the residual error goes to zero is the true `nec`, which is why
the precision axis is the only way to read it.

The arms that keep the measurement behave as that definition predicts. `measured`
falls from +13.8% to +3.9%, `censored` from +13.4% to +2.1% and `deleted` from
+9.5% to +1.2%, all approaching the truth from above.

The arms that replace it go the other way and keep going. `floored` moves from
-10.5% to -52.9%, `beta` from -6.8% to -43.2%, and `gamma` from +10.1% to
-69.5%. Improving the experiment makes these three worse, and at the highest
precision tested they place the no-effect concentration between two and three
times too low.

## Interval coverage

Bias understates the problem, because the intervals narrow around a displaced
centre as the experiment improves.

```{r fig-coverage, fig.height=4.2}
#| eval: !expr have_sim
d <- filter(prec, estimate == "ErC50")
ggplot(d, aes(cv, coverage, colour = arm, group = arm)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "coverage") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Proportion of 95% intervals containing the true ErC50") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 4.** Coverage of the 95% credible interval for ErC50. The dashed line
is nominal coverage. Monte Carlo standard error is about 1.5 percentage points
at 100 iterations.

```{r tab-coverage}
#| eval: !expr have_sim
m |>
  filter(estimate == "ErC50", cell %in% c("p1", "p4")) |>
  mutate(cell = ifelse(cell == "p1", "CV 9.6%", "CV 1.9%")) |>
  select(cell, arm, rel_bias, coverage, width, rmse) |>
  arrange(cell, arm) |>
  knitr::kable(digits = c(0, 0, 2, 2, 3, 3),
    col.names = c("cell", "arm", "bias (%)", "coverage", "mean width", "RMSE"),
    caption = "**Table 2.** ErC50 at the two ends of the precision axis. Coverage and interval width are reported together, because either alone identifies the wrong arm as the best one.")
```

At the highest precision tested, `floored` contains the true ErC50 in 24 of 100
intervals and `beta` in 48, against a nominal 95. Their bias at that point is
only -2.3% and -6.0%; it is the narrowing of the interval around a displaced
centre that destroys the coverage.

A laboratory that improves its technique, adds replicates and reduces its
control variability will, if it replaces its negative values, obtain an analysis
that is more confident and no more accurate.

Coverage and width have to be read together. `gamma` reaches 0.94 coverage at
the highest precision, which looks respectable until its mean interval width of
0.296 is compared with `measured`'s 0.134, and its ErC10 bias of -24% is
recalled. Either statistic alone identifies the wrong arm as the best one.

## Where the response barely goes negative

```{r tab-ctl}
#| eval: !expr have_sim
m |>
  filter(cell == "ctl", estimate %in% c("ErC50", "NSEC")) |>
  select(estimate, arm, rel_bias, coverage) |>
  arrange(estimate, arm) |>
  knitr::kable(digits = c(0, 0, 2, 2),
    col.names = c("estimate", "arm", "bias (%)", "coverage"),
    caption = "**Table 3.** The control cell, in which the concentration series stops where the true curve reaches zero growth, so that almost no true mean is negative.")
```

Where the design does not reach the zero-growth boundary, the four Gaussian arms
agree to within about two per cent on ErC50 and the conventions have nothing to
act on. The two family-imposed arms still differ, because scaling to a
proportion or fitting a Gamma changes the analysis whether or not any value is
negative.

This is the condition under which none of the results above apply. Reporting the
proportion of observations at or below zero is what tells a reader which
situation they are in.

## Convergence, read against model weight

Every fit records `rhat` and the tail effective sample size for the curve
parameters of each equation in its candidate set, beside the stacking weight
that equation holds. The two are only readable together. An equation whose shape
suits the data badly fails a diagnostic whatever the data are doing, and is
given almost no weight for the same reason, so a failure on a near-zero-weight
equation describes that curve rather than the model-averaged estimate.

```{r tab-diagnostics}
#| eval: !expr (!is.null(diagnostics))
diagnostics |>
  group_by(cell, arm) |>
  summarise(w_rhat = mean(tapply(weight * (max_rhat > 1.01),
                                 interaction(iteration), sum)),
            w_ess  = mean(tapply(weight * (min_ess_tail < 400),
                                 interaction(iteration), sum)),
            .groups = "drop") |>
  arrange(desc(w_rhat + w_ess)) |>
  head(12) |>
  knitr::kable(digits = 3,
    col.names = c("cell", "arm", "weight on rhat > 1.01",
                  "weight on tail ESS < 400"),
    caption = "**Table 4.** Mean stacking weight held by equations that fail a convergence diagnostic, by cell and arm, over the realisations of that cell. Twelve worst shown. A row near zero means the diagnostics that failed were on equations the average barely used.")
```

The previous run of this study recorded no diagnostics at all, so when the
question arose of whether a badly mixing equation had affected an estimate there
was nothing on disk to answer it with. That is why they are collected here.

## The case studies

Four marine microalgal growth tests were analysed by the same six conventions,
from the `alga` data that ship with `bayesnec`: two tests on *Cladocopium
proliferum* over seven days and two on *Rhodomonas salina* over three, each
species against two contaminants on undisclosed and mutually incomparable
concentration scales.

They are run here rather than when the vignette is built. Four datasets by six
conventions, each averaged over thirteen or fourteen equations, is a fifteen-hour
build, which is not something to run on a workstation whenever a sentence
changes. What that costs is stated rather than hidden: the case-study estimates
no longer track whatever version of `bayesnec` they are published beside. They
report the commit named in `hpc/bayesnec.lock`, exactly as the simulation does.

The two *Rhodomonas* tests need one qualification before their results are read.
At the highest concentrations the population fell below the counting limit of 10
cells per mL, so the growth rate is undefined rather than negative. Every
convention must therefore substitute something there, including the ones that
otherwise retain the measurement, and on those two datasets no convention is a
true analysis of what was measured. Censoring is the exception in kind rather
than in degree: it can state that the value lies at or below the rate the
counting limit implies without inventing one.

```{r tab-cases}
#| eval: !expr have_cases
cases |>
  filter(estimate == "ErC50") |>
  select(dataset, arm, value, lower, upper, top_model, top_weight,
         w_zero_asymptote) |>
  arrange(dataset, arm) |>
  knitr::kable(digits = c(0, 0, 3, 3, 3, 0, 2, 2),
    col.names = c("dataset", "arm", "ErC50", "lower", "upper",
                  "highest-weighted", "its weight", "weight on zero-asymptote"),
    caption = "**Table 5.** ErC50 from each convention on each dataset, with the equation the average leaned on most and the combined weight on the six equations whose lower asymptote is fixed at zero. There is no truth to score these against; what they show is how far the choice of convention changes the number that would be reported.")
```

```{r fig-case-ratio, fig.height=5}
#| eval: !expr have_cases
# Ratios rather than values: the four tests are on undisclosed and mutually
# incomparable concentration scales, so only the within-dataset comparison
# means anything.
cr <- cases |>
  group_by(dataset, estimate) |>
  mutate(ref = value[arm == "measured"],
         r = value / ref, rlo = lower / ref, rhi = upper / ref) |>
  ungroup()
ggplot(cr, aes(x = r, y = arm, colour = arm)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_errorbar(aes(xmin = rlo, xmax = rhi), orientation = "y",
                width = 0, linewidth = 0.6) +
  geom_point(size = 2.2) +
  scale_y_discrete(limits = rev(ARMS)) +
  scale_x_log10() +
  facet_grid(estimate ~ dataset) +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  labs(x = "estimate as a ratio to the `measured` analysis (log scale)",
       y = NULL) +
  theme_classic(base_size = 9) +
  theme(strip.background = element_blank())
```

## What the study establishes

The measurement should be kept. Across the precision axis `measured` recovers
the true ErC50 to within `r f1(max(abs(vapply(c("p1","p2","p3","p4"), function(cl) val("measured","ErC50",cl), numeric(1)))))`%,
and at the highest precision tested its interval contains the truth in
`r f2(val("measured","ErC50","p4","coverage"))` of realisations.

Where the negative values were never recorded, censoring recovers almost all of
what replacing them loses: at the highest precision `censored` reports an ErC50
bias of `r f1(val("censored","ErC50","p4"))`% against
`r f1(val("measured","ErC50","p4"))`% for the measurements themselves.

Deleting the negative rows is better than replacing them with zero and worse
than either alternative above. It discards information that censoring retains,
and it changes the replication per concentration rather than the values, which
at the top of a series can remove a treatment group altogether.

Replacing negative values with zero leaves an ErC50 bias of
`r f1(val("floored","ErC50","p4"))`% at the highest precision that does not
shrink as the experiment improves, so interval coverage falls to
`r f2(val("floored","ErC50","p4","coverage"))`, and it displaces the NSEC by
`r f1(val("floored","NSEC","p4"))`%.

Imposing the boundary through the response distribution is the same practice
reached by a different route and performs no better. `beta` holds an ErC50 bias
of `r f1(val("beta","ErC50","p4"))`% at the highest precision;
`gamma` is close to unbiased on ErC50 and displaces ErC10 by
`r f1(val("gamma","ErC10","p4"))`%, with coverage of
`r f2(val("gamma","ErC10","p4","coverage"))`, so the two estimates cannot be
summarised together for that convention.

Where the design does not reach the zero-growth boundary none of this applies:
in the `ctl` cell the four Gaussian conventions span
`r f1(spread("ErC50","ctl"))` percentage points of ErC50 bias between them,
against `r f1(spread("ErC50","p1"))` in `p1`, which has the same control
variability and four times as many negative observations.

## Provenance

The study is 4,200 model-averaged fits for the simulation — seven cells, six
conventions, 100 realisations — and 24 for the case studies, each fit averaging
over the declining candidate set its family allows. It ran on the AIMS HPC as
SLURM arrays, one unit per task.

`bayesnec` is pinned at the commit named in `hpc/bayesnec.lock`, on the `dev`
branch. That is a development commit rather than a CRAN release, and the
difference matters here more than it usually would, because what this study
measures is what package defaults do.

This is a re-run. The previous results, produced at `0976caca`, are kept under
`archive/bayesnec-0976caca/` and cannot be pooled with these. Three changes
between the two alter every fit:

- the `nec` and `ec50` priors are now built on the log of the predictor, so
  their central 95% interval covers the concentrations tested rather than a
  fraction of them. On the reference cell the `nec` prior went from
  `gamma(5, 2.666)` to `lognormal(0.615, 1.175)`, which moves the true ErC50
  from above the 99th percentile of its own prior to the 80th;
- the regularizing prior set was redefined and anchored on the predictor. This
  study uses the default `uninformative` set, so that changes nothing here
  directly and is named because it is part of the same release;
- the initial-value search was rewritten to accept chains individually and to
  band the curve on level means, which changes the starting values of every fit.

The container holds the toolchain and not the package: it is the image built by
the `bayesnec` repository for precompiling its vignettes, and the job installs
`bayesnec` from the pinned source at start-up. `hpc/image.lock` records which
image produced these results and every job refuses to run against one whose
SHA does not match it. Rebuilding the image is a change that can alter published
numbers, so it is made deliberately rather than discovered afterwards.

The simulation's priors are held fixed within a cell and arm rather than derived
from each realisation. `bayesnec` derives its defaults from the response vector
and `brms` writes them into the Stan source as literals, so a prior that varies
by realisation makes every fit a distinct Stan program: 5,622 compiles for 406
units before this was changed. Fixing them makes the program identical across
iterations. Each arm keeps its own prior, derived from its own prepared data, so
the contrast between arms is unaffected, and the exact values used are committed
as 42 files in `priors/`.

The case studies take `bnec()`'s own defaults instead. Each dataset and arm is
fitted once, so there is no repetition for an identical program to save, and
taking the defaults is the practice they exist to show.

The generating model is carried over verbatim from
[open-AIMS/negative-sgr](https://github.com/open-AIMS/negative-sgr) at commit
`0181d66e`. Nothing else is.

```{r tab-provenance}
lock <- readLines(here::here("hpc", "bayesnec.lock"))
img <- readLines(here::here("hpc", "image.lock"))
get1 <- function(x, k) sub(paste0("^", k, ": "), "",
                           grep(paste0("^", k, ": "), x, value = TRUE)[1])
knitr::kable(
  data.frame(
    item = c("bayesnec commit", "bayesnec version", "branch",
             "base image", "R", "cmdstan", "CRAN snapshot", "image sha256"),
    value = c(get1(lock, "commit"), get1(lock, "version"), get1(lock, "branch"),
              get1(img, "base_image"), get1(img, "r_version"),
              get1(img, "cmdstan"), get1(img, "repos"),
              substr(get1(img, "sif_sha256"), 1, 16)),
    check.names = FALSE),
  caption = "**Table 6.** What produced these numbers, read from the two lock files rather than typed.")
```
