---
title: "Conventions for handling negative response values: simulation results"
format:
  html:
    toc: true
    embed-resources: true
    fig-width: 9
    fig-height: 5
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(ggplot2); library(dplyr); library(tidyr)
m <- read.csv(here::here("results", "metrics_100iter.csv"))

ARMS <- c("measured", "censored", "deleted", "floored", "beta", "gamma")
# Hue order validated for colour-vision separation as a categorical set: every
# adjacent pair clears the CVD and normal-vision floors on a light surface.
# Reordering the arms means revalidating, because the check is on adjacent pairs.
ARM_COL <- c(measured = "#1B6BC4", censored = "#0E7C61", deleted  = "#C25E12",
             floored  = "#B0217A", beta     = "#7C7A00", gamma    = "#6B4EBF")
m$arm <- factor(m$arm, levels = ARMS)
m$estimate <- factor(m$estimate, levels = c("ErC10", "ErC50", "NSEC"))

prec <- m |>
  filter(cell %in% c("p1", "p2", "p3", "p4")) |>
  mutate(cv = factor(sprintf("%.1f", 100 * cv_control),
                     levels = c("9.6", "6.7", "2.8", "1.9")))

# One label per series at the right-hand edge instead of a key: six hues in a
# legend are harder to match to lines than a label sitting on the line it names.
edge_label <- function(d, yvar) {
  geom_text(data = subset(d, cv == "1.9"),
            aes(label = arm, y = .data[[yvar]]), hjust = -0.15, size = 2.6,
            show.legend = FALSE)
}
```

## The study

We simulated concentration-response data from a four-parameter curve
(`nec4param`) in which specific growth rate falls from a positive control value
to a negative lower asymptote, and we handed each simulated dataset to six
analyses that differ only in how they treat the negative values. Every analysis
uses `bayesnec` with package defaults, model-averaged over the declining
candidate set the family allows, so a difference between arms is the difference
a practitioner would obtain rather than a likelihood-only contrast.

```{r arms-table}
knitr::kable(
  data.frame(
    arm = ARMS,
    `what the analyst does` = c(
      "use the values as recorded",
      "declare negatives left-censored at zero",
      "delete the negative rows, keep the rest of their group",
      "replace negatives with zero",
      "floor, scale to (0, 1), fit a Beta",
      "floor, fit a Gamma"),
    family = c("gaussian", "gaussian", "gaussian", "gaussian", "Beta", "Gamma"),
    check.names = FALSE),
  caption = "**Table 1.** The six conventions compared. All six are reachable from `bnec()` defaults.")
```

Seven design cells were run, each 100 times. Four vary measurement precision
alone, holding the design and the true values fixed, so that the control
coefficient of variation falls from 9.6% to 1.9%. Two vary the depth of the
lower asymptote. One stops the concentration series at the point where the true
curve reaches zero growth, so that almost nothing is negative and the
conventions have nothing to act on.

The true values are known by construction, so each analysis is scored against
them rather than against each other. We report relative bias, the proportion of
95% credible intervals containing the truth, and the mean interval width.

## Estimating ErC50

The concentration halving the growth rate separates the conventions cleanly, and
it separates them by whether the measurement is kept.

```{r fig-erc50-bias, fig.height=4.2}
d <- filter(prec, estimate == "ErC50")
ggplot(d, aes(cv, rel_bias, colour = arm, group = arm)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "rel_bias") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Relative bias in ErC50 (%)") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 1.** Relative bias in the estimated ErC50 against measurement
precision, 100 simulated datasets per point. Arms that keep the measurement
converge on the true value; arms that replace it level off away from zero.

Three arms converge as the experiment improves: `measured` from -0.8% to -0.2%,
`censored` from -1.3% to -0.2%, and `deleted` from -2.5% to 0.0%. Their error is
estimation error, and it behaves like estimation error.

Two do not. `floored` settles at about -2.3% and `beta` at about -6.0%, and
neither improves as the control coefficient of variation falls from 9.6% to
1.9%. A discrepancy that does not shrink as the noise shrinks is
misspecification, not imprecision, and no amount of experimental effort removes
it.

`gamma` is a special case that the bias alone misrepresents. Its ErC50 bias is
near zero in every cell, but its intervals are more than twice as wide as any
other arm's and its ErC10 is badly displaced, as the next section shows.

## Estimating ErC10

```{r fig-erc10-bias, fig.height=4.2}
d <- filter(prec, estimate == "ErC10")
ggplot(d, aes(cv, rel_bias, colour = arm, group = arm)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "rel_bias") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Relative bias in ErC10 (%)") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 2.** Relative bias in the estimated ErC10. `gamma` is displaced by
about 20% at every level of precision.

The four Gaussian arms are within a few per cent of the truth once the
measurement is precise enough to resolve a 10% decline, and they converge.
`gamma` does not: it underestimates ErC10 by 19% to 24% in every cell of the
precision axis, and the error is flat, so a better experiment does not help.
`beta` drifts the other way, from -4.8% to +8.9%.

ErC10 is the harder estimate for every arm at realistic noise, because a 10%
decline from the control has to be resolved against a control coefficient of
variation of 9.6%. That is a property of the endpoint rather than of any
convention, and it is why the precision axis is necessary to interpret it.

## The NSEC

The no-significant-effect concentration is where the conventions diverge most,
and where they diverge in a direction that matters for protection.

```{r fig-nsec-bias, fig.height=4.2}
d <- filter(prec, estimate == "NSEC")
ggplot(d, aes(cv, rel_bias, colour = arm, group = arm)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "rel_bias") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Relative bias in the NSEC (%)") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 3.** Relative bias in the estimated NSEC against the true no-effect
concentration. The three arms that keep the measurement converge toward it from
above; the three that replace it diverge downward as precision improves.

The NSEC's expected value exceeds the true `nec` at any finite level of noise,
because it is defined against a lower quantile of the control posterior: a
noisier control gives a lower reference and pushes the crossing to the right. Its
limiting value as the residual error goes to zero is the true `nec`, which is why
the precision axis is the only way to read it.

The arms that keep the measurement behave as that definition predicts. `measured`
falls from +13.8% to +3.9%, `censored` from +13.4% to +2.1% and `deleted` from
+9.5% to +1.2%, all approaching the truth from above.

The arms that replace it go the other way and keep going. `floored` moves from
-10.5% to -52.9%, `beta` from -6.8% to -43.2%, and `gamma` from +10.1% to
-69.5%. Improving the experiment makes these three worse, and at the highest
precision tested they place the no-effect concentration between two and three
times too low.

## Interval coverage

Bias understates the problem, because the intervals narrow around a displaced
centre as the experiment improves.

```{r fig-coverage, fig.height=4.2}
d <- filter(prec, estimate == "ErC50")
ggplot(d, aes(cv, coverage, colour = arm, group = arm)) +
  geom_hline(yintercept = 0.95, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.9) +
  edge_label(d, "coverage") +
  scale_colour_manual(values = ARM_COL, guide = "none") +
  scale_x_discrete(expand = expansion(add = c(0.3, 1.1))) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Control coefficient of variation (%) — more precise →",
       y = "Proportion of 95% intervals containing the true ErC50") +
  theme_classic(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey92"))
```

**Figure 4.** Coverage of the 95% credible interval for ErC50. The dashed line
is nominal coverage. Monte Carlo standard error is about 1.5 percentage points
at 100 iterations.

```{r tab-coverage}
m |>
  filter(estimate == "ErC50", cell %in% c("p1", "p4")) |>
  mutate(cell = ifelse(cell == "p1", "CV 9.6%", "CV 1.9%")) |>
  select(cell, arm, rel_bias, coverage, width, rmse) |>
  arrange(cell, arm) |>
  knitr::kable(digits = c(0, 0, 2, 2, 3, 3),
    col.names = c("cell", "arm", "bias (%)", "coverage", "mean width", "RMSE"),
    caption = "**Table 2.** ErC50 at the two ends of the precision axis. Coverage and interval width are reported together, because either alone identifies the wrong arm as the best one.")
```

At the highest precision tested, `floored` contains the true ErC50 in 24 of 100
intervals and `beta` in 48, against a nominal 95. Their bias at that point is
only -2.3% and -6.0%; it is the narrowing of the interval around a displaced
centre that destroys the coverage.

A laboratory that improves its technique, adds replicates and reduces its
control variability will, if it replaces its negative values, obtain an analysis
that is more confident and no more accurate.

Coverage and width have to be read together. `gamma` reaches 0.94 coverage at
the highest precision, which looks respectable until its mean interval width of
0.296 is compared with `measured`'s 0.134, and its ErC10 bias of -24% is
recalled. Either statistic alone identifies the wrong arm as the best one.

## Where the response barely goes negative

```{r tab-ctl}
m |>
  filter(cell == "ctl", estimate %in% c("ErC50", "NSEC")) |>
  select(estimate, arm, rel_bias, coverage) |>
  arrange(estimate, arm) |>
  knitr::kable(digits = c(0, 0, 2, 2),
    col.names = c("estimate", "arm", "bias (%)", "coverage"),
    caption = "**Table 3.** The control cell, in which the concentration series stops where the true curve reaches zero growth, so that almost no true mean is negative.")
```

Where the design does not reach the zero-growth boundary, the four Gaussian arms
agree to within about two per cent on ErC50 and the conventions have nothing to
act on. The two family-imposed arms still differ, because scaling to a
proportion or fitting a Gamma changes the analysis whether or not any value is
negative.

This is the condition under which none of the results above apply. Reporting the
proportion of observations at or below zero is what tells a reader which
situation they are in.

## What the study establishes

The measurement should be kept. `measured` is unbiased to within one per cent on
ErC50 at every level of precision tested, converges on the truth for all three
estimates, and holds nominal coverage.

Where the negative values were never recorded, censoring recovers almost all of
what replacing them loses: `censored` tracks `measured` to within about half a
percentage point on ErC50 across the precision axis.

Deleting the negative rows is better than replacing them with zero but worse
than either alternative above, and it discards information that censoring
retains.

Replacing negative values with zero biases ErC50 low, places the NSEC two to
three times too low at high precision, and drives interval coverage to 0.24. The
error does not shrink as the experiment improves.

Imposing the boundary through the response distribution is the same practice
reached by a different route, and it performs no better: `beta` carries the
largest ErC50 bias in the study and `gamma` displaces ErC10 by about 20% at
every level of precision while producing the widest intervals of any arm.

## Provenance

The study is 4,200 model-averaged fits: seven cells, six arms, 100 iterations,
each fit averaging over the declining candidate set for its family. It ran on
the AIMS HPC as a SLURM array, one unit per task, with zero failures. Peak
memory was 1.93 GB per unit and the mean unit took 936 seconds.

`bayesnec` was pinned at commit `0976caca`, version 2.1.3.33, from the `dev`
branch. That is not the CRAN release: the current CRAN version is 2.1.3.1, which
does not include the change peaking the `nec` and `ec50` gamma prior at the
median predictor rather than at twice it. The priors are therefore a development
default rather than a released one, and the exact values used are committed as
42 files in `priors/`, one per cell and arm.

Those priors are held fixed within a cell and arm rather than being derived from
each realisation. `bayesnec` derives its defaults from the response vector and
`brms` writes them into the Stan source as literals, so a prior that varies by
realisation makes every fit a distinct Stan program: 5,622 compiles for 406
units before this was changed. Fixing them makes the program identical across
iterations. The predictor range is identical across realisations within a cell
and arm, so the hard parameter bounds do not vary at all; only the
response-derived scales would have, by a few per cent. Each arm keeps its own
prior, derived from its own prepared data, so the contrast between arms is
unaffected.

The generating model is carried over verbatim from
[open-AIMS/negative-sgr](https://github.com/open-AIMS/negative-sgr) at commit
`0181d66e`. Nothing else is.
